### 2019-10-16 ch3_dynbenchmark/content.tex
change cites

\section{Introduction}
Single-cell omics data, including transcriptomics, proteomics and epigenomics data, provide new opportunities for studying cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation \cite{tanay_scalingsinglecellgenomics_2017,etzrodt_quantitativesinglecellapproaches_2014}. Such dynamic processes can be modeled computationally using trajectory inference (TI) methods, also called pseudotime analysis, which order cells along a trajectory based on similarities in their expression patterns \cite{trapnell_definingcelltypes_2015,cannoodt_computationalmethodstrajectory_2016,moon_manifoldlearningbasedmethods_2018}. The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also identify more complex trajectory topologies, such as cyclic \cite{liu_reconstructingcellcycle_2017} or disconnected graphs \cite{wolf_graphabstractionreconciles_2017}. TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process\cite{tanay_scalingsinglecellgenomics_2017}, thereby allowing the objective identification of new (primed) subsets of cells \cite{schlitzer_identificationcdc1cdc2committed_2015}, delineation of a differentiation tree \cite{velten_humanhaematopoieticstem_2017,see_mappinghumandc_2017} and inference of regulatory interactions responsible for one or more bifurcations \cite{aibar_scenicsinglecellregulatory_2017}. Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogs of whole organisms \cite{regev_humancellatlas_2017,han_mappingmousecell_2018,schaum_singlecelltranscriptomics20_2018} underline the urgency for accurate, scalable \cite{aibar_scenicsinglecellregulatory_2017,angerer_singlecellsmake_2017} and user-friendly TI methods.

>>>>

\section{Introduction}
Single-cell omics data, including transcriptomics, proteomics and epigenomics data, provide new opportunities for studying cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation \cite{tanay_scalingsinglecellgenomics_2017,etzrodt_quantitativesinglecellapproaches_2014}. Such dynamic processes can be modeled computationally using trajectory inference (TI) methods, also called pseudotime analysis, which order cells along a trajectory based on similarities in their expression patterns \cite{trapnell_definingcelltypes_2015,cannoodt_computationalmethodstrajectory_2016,moon_manifoldlearningbasedmethods_2018}. The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also identify more complex trajectory topologies, such as cyclic \cite{liu_reconstructingcellcycle_2017} or disconnected graphs \cite{wolf_graphabstractionreconciles_2017}. TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process \cite{tanay_scalingsinglecellgenomics_2017}, thereby allowing the objective identification of new (primed) subsets of cells \cite{schlitzer_identificationcdc1cdc2committed_2015}, delineation of a differentiation tree \cite{velten_humanhaematopoieticstem_2017,see_mappinghumandc_2017} and inference of regulatory interactions responsible for one or more bifurcations \cite{aibar_scenicsinglecellregulatory_2017}. Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogs of whole organisms \cite{regev_humancellatlas_2017,han_mappingmousecell_2018,schaum_singlecelltranscriptomics20_2018} underline the urgency for accurate, scalable \cite{aibar_scenicsinglecellregulatory_2017,angerer_singlecellsmake_2017} and user-friendly TI methods.

### 2019-10-16 ch6_bred/content.tex
change cites

One of the central cellular processes underlying development is transcriptional regulation. During development, changes in transcription factor activity induce chromatin modifications, chromatin remodelling and ultimately a differential recruitment of the basal transcriptional machinery\cite{coulon_eukaryotictranscriptionaldynamics_2013}. Modelling the dynamics of gene regulation is therefore essential to better understand why a cellular dynamic processes progresses through several steps, and what goes wrong in the case of disease.

The dynamics of gene regulation has classically been studied using time series data\cite{bar-joseph_studyingmodellingdynamic_2012}. When dynamic processes progress asynchronously, such as in hematopoiesis, time series data are usually obtained by sorting different transition states and assessing bulk gene expression and transcription factor binding within the population\cite{novershtern_denselyinterconnectedtranscriptional_2011, may_dynamicanalysisgene_2013, jojic_identificationtranscriptionalregulators_2013, goode_dynamicgeneregulatory_2016}. Alternatively, time series data can also be generated by synchronizing the dynamic process between cells. However, issues with time-resolution, heterogeneity and good in vivo synchronization models can often limit the predictive power of the dynamic models of gene regulation which can be constructed\cite{bar-joseph_studyingmodellingdynamic_2012}.

@@

Several studies have highlighted how some regulatory interactions can be very dynamic while others show evidence of being static during consecutive developmental stages\cite{moignard_characterizationtranscriptionalnetworks_2013, pina_singlecellnetworkanalysis_2015}. 
Since regulatory interactions are context-dependent\cite{papp_genomewideanalysiscontextdependence_2005}, attempting to create an accurate model of those processes by inferring a static regulatory network may have limited relevance.
Case-wise NI methods\footnote{Case-wise NI is sometimes also called sample-specific NI or case-specific NI.} avoid predicting a static GRN and instead infer one GRN per cell (or per sample, for bulk omics data).

In order to compute a case-wise GRN for a single sample, Kuijjer et al.\cite{kuijjer_estimatingsamplespecificregulatory_2019} and Liu et al.\cite{liu_personalizedcharacterizationdiseases_2016} employ similar strategies, namely by computing the difference of computing a static GRN for all the cases, and computing a static GRN for all the cases minus one. Since this procedure needs to be repeated for every case in the dataset, and because NI methods are already amongst the most computationally intensive analyses to perform on omics data, this methodology is not applicable for large omics datasets.
Another case-wise NI method, SCENIC\cite{aibar_scenicsinglecellregulatory_2017} infers case-wise GRNs by first inferring a static GRN using GENIE3\cite{huynh-thu_inferringregulatorynetworks_2010}. GENIE3 is a static NI method which uses Random Forests\cite{breiman_randomforests_2001} feature importance scores to prioritise candidate regulators for a particular target gene. SCENIC then post-processes the static GRN to determine whether an interaction is enriched for particular cases, resulting in a case-wise GRN. 
In short, while several case-wise NI methods thus already exist, their implementation consisted of post-processing a static GRN to arrive at a case-wise GRN.

In this work, we introduce \texttt{bred}, the first 'true' case-wise NI method. For each interaction in the inferred case-wise GRN, \texttt{bred} predicts both the regulatory strength and its effect. 
The case-wise GRNs -- or 'case-wise regulomes' -- can be analysed analogously to transcriptomics data; for example by clustering samples, inferring trajectories, or finding differentially activated regions.
We demonstrate \texttt{bred} by applying it to a single-cell dataset of 22'122 hematopoietic cells from the Tabula Muris project\cite{schaum_singlecelltranscriptomics20_2018}, and to a collection of 14'963 bulk omics samples from The Cancer Genome Atlas project\cite{weinstein_cancergenomeatlas_2013}.

@@

\subsection{Hematopoietic cells from Tabula Muris}
From the Tabula Muris\cite{schaum_singlecelltranscriptomics20_2018} project, we selected all cells involved in some part of the hematopoietic lineage tree. We computed case-wise GRNs between 22'122 remaining single cells, for the 2000 most variable target genes and a subset of transcription factors as regulators.

To summarise the similarity in GRNs across cases, we first visualise a dimensionality reduction of the case-wise GRNs, where every dot represents a single GRN (Figure~\ref{fig:tabmur}A). The dimensionality reduction was computed by applying Fruchterman-Reingold\cite{fruchterman_graphdrawingforcedirected_1991} on the $k$-nearest-neighbour ($k$NN) graph of the case-wise GRN vectors. On the same $k$NN graph, the cells were clustered using the Louvain\cite{blondel_fastunfoldingcommunities_2008} clustering algorithm, and the clusters were labelled using prior information from Tabula Muris.
For each cluster, we retained the 50 interactions with the highest mean importance scores (Figure~\ref{fig:tabmur}B). In the visualisation, each node represents a gene, each edge an interaction, and the colour represents which cluster this interaction belongs to.

@@

\subsection{Inferring case-wise GRNs}
The task of inferring a static GRN (Figure~\ref{fig:simplify}A) can be reduced to a simpler problem, namely: for every target $T$, predict which of the potential regulators $R_i$ regulate $T$ (Figure~\ref{fig:simplify}B). This simplification allowed GENIE3\cite{huynh-thu_inferringregulatorynetworks_2010} to use Random Forest's\cite{breiman_randomforests_2001} feature importance scores for inferring GRNs. Namely, a Random Forest is trained to predict the expression of a target gene of interest from the expression of potential regulators. The resulting Random Forest inherently allows to extract a feature importance score by observing the effect of each regulator in making a good prediction for the target expression. As in GENIE3, the target expression is first scaled to normalise feature importance scores across different targets.

@@

Computing the case-wise feature importances of a tree consists of the following 8 steps (Figure~\ref{fig:fimp}).
The 'randomness' of a Random Forest is due to only using a subset of the samples in the dataset in order to build a single decision tree. The samples are split into two groups, the 'in-bag' data and the 'out-of-bag' data (Figure~\ref{fig:fimp}A). A decision tree\cite{breiman_classificationregressiontrees_1984} is trained on the in-bag expression of the potential regulators in trying to predict the in-bag target gene expression (Figure~\ref{fig:fimp}B). The target expression of the out-of-bag samples is predicted using the decision tree (Figure~\ref{fig:fimp}C), and the squared error between the real and target expression is computed (Figure~\ref{fig:fimp}D). For each sample in the out-of-bag set, this vector represents how well the decision tree was able to predict the expression of the target gene.

@@

To compute the case-wise GRNs, we implemented the abovementioned methodology in C++ in a modified version of the \texttt{ranger} R/C++ package\cite{wright_rangerfastimplementation_2017}.

@@

\subsection{Clustering of case-wise GRNs}
To perform downstream analysis on the cases, first a $k$-nearest neighbour ($k$NN) graph of the cases is computed.
In order for the $k$NN graph to better emphasise similarities in GRNs rather than absolute euclidean distances, we first reduce the dimensionality of the case-by-interaction matrix to case-by-20 matrix using Landmark Multi-Dimensional Scaling\cite{lee_landmarkmdsensemble_2009} with a Spearman rank distance metric.

Next, KD-trees are used to calculate the $k$NN graph efficiently. The cases in the dataset are visualised and clusted using the Fruchterman-Reingold\cite{fruchterman_graphdrawingforcedirected_1991} and Louvain clustering\cite{blondel_fastunfoldingcommunities_2008}, respectively.

The following R packages provided implementations for each of these algorithms: lmds, RANN, igraph\cite{csardi_igraphsoftwarepackage_2006}.

\subsection{Visualising clustered GRNs}
After Louvain clustering, the interactions of the 50 interactions with highest mean importance per cluster are retained. These interactions are visualised in Cytoscape\cite{shannon_cytoscapesoftwareenvironment_2003}, in which nodes depict genes, edges depict predicted regulatory interactions, coloured according to which cluster they are predicted for. The shape of the arrow denotes the predicted effect of the regulatory interaction.

>>>>

One of the central cellular processes underlying development is transcriptional regulation. During development, changes in transcription factor activity induce chromatin modifications, chromatin remodelling and ultimately a differential recruitment of the basal transcriptional machinery \cite{coulon_eukaryotictranscriptionaldynamics_2013}. Modelling the dynamics of gene regulation is therefore essential to better understand why a cellular dynamic processes progresses through several steps, and what goes wrong in the case of disease.

The dynamics of gene regulation has classically been studied using time series data \cite{bar-joseph_studyingmodellingdynamic_2012}. When dynamic processes progress asynchronously, such as in hematopoiesis, time series data are usually obtained by sorting different transition states and assessing bulk gene expression and transcription factor binding within the population \cite{novershtern_denselyinterconnectedtranscriptional_2011, may_dynamicanalysisgene_2013, jojic_identificationtranscriptionalregulators_2013, goode_dynamicgeneregulatory_2016}. Alternatively, time series data can also be generated by synchronizing the dynamic process between cells. However, issues with time-resolution, heterogeneity and good in vivo synchronization models can often limit the predictive power of the dynamic models of gene regulation which can be constructed \cite{bar-joseph_studyingmodellingdynamic_2012}.

@@

Several studies have highlighted how some regulatory interactions can be very dynamic while others show evidence of being static during consecutive developmental stages \cite{moignard_characterizationtranscriptionalnetworks_2013, pina_singlecellnetworkanalysis_2015}. 
Since regulatory interactions are context-dependent \cite{papp_genomewideanalysiscontextdependence_2005}, attempting to create an accurate model of those processes by inferring a static regulatory network may have limited relevance.
Case-wise NI methods\footnote{Case-wise NI is sometimes also called sample-specific NI or case-specific NI.} avoid predicting a static GRN and instead infer one GRN per cell (or per sample, for bulk omics data).

In order to compute a case-wise GRN for a single sample, Kuijjer et al. \cite{kuijjer_estimatingsamplespecificregulatory_2019} and Liu et al. \cite{liu_personalizedcharacterizationdiseases_2016} employ similar strategies, namely by computing the difference of computing a static GRN for all the cases, and computing a static GRN for all the cases minus one. Since this procedure needs to be repeated for every case in the dataset, and because NI methods are already amongst the most computationally intensive analyses to perform on omics data, this methodology is not applicable for large omics datasets.
Another case-wise NI method, SCENIC \cite{aibar_scenicsinglecellregulatory_2017} infers case-wise GRNs by first inferring a static GRN using GENIE3 \cite{huynh-thu_inferringregulatorynetworks_2010}. GENIE3 is a static NI method which uses Random Forests \cite{breiman_randomforests_2001} feature importance scores to prioritise candidate regulators for a particular target gene. SCENIC then post-processes the static GRN to determine whether an interaction is enriched for particular cases, resulting in a case-wise GRN. 
In short, while several case-wise NI methods thus already exist, their implementation consisted of post-processing a static GRN to arrive at a case-wise GRN.

In this work, we introduce \texttt{bred}, the first 'true' case-wise NI method. For each interaction in the inferred case-wise GRN, \texttt{bred} predicts both the regulatory strength and its effect. 
The case-wise GRNs -- or 'case-wise regulomes' -- can be analysed analogously to transcriptomics data; for example by clustering samples, inferring trajectories, or finding differentially activated regions.
We demonstrate \texttt{bred} by applying it to a single-cell dataset of 22'122 hematopoietic cells from the Tabula Muris project \cite{schaum_singlecelltranscriptomics20_2018}, and to a collection of 14'963 bulk omics samples from The Cancer Genome Atlas project \cite{weinstein_cancergenomeatlas_2013}.

@@

\subsection{Hematopoietic cells from Tabula Muris}
From the Tabula Muris \cite{schaum_singlecelltranscriptomics20_2018} project, we selected all cells involved in some part of the hematopoietic lineage tree. We computed case-wise GRNs between 22'122 remaining single cells, for the 2000 most variable target genes and a subset of transcription factors as regulators.

To summarise the similarity in GRNs across cases, we first visualise a dimensionality reduction of the case-wise GRNs, where every dot represents a single GRN (Figure~\ref{fig:tabmur}A). The dimensionality reduction was computed by applying Fruchterman-Reingold \cite{fruchterman_graphdrawingforcedirected_1991} on the $k$-nearest-neighbour ($k$NN) graph of the case-wise GRN vectors. On the same $k$NN graph, the cells were clustered using the Louvain \cite{blondel_fastunfoldingcommunities_2008} clustering algorithm, and the clusters were labelled using prior information from Tabula Muris.
For each cluster, we retained the 50 interactions with the highest mean importance scores (Figure~\ref{fig:tabmur}B). In the visualisation, each node represents a gene, each edge an interaction, and the colour represents which cluster this interaction belongs to.

@@

\subsection{Inferring case-wise GRNs}
The task of inferring a static GRN (Figure~\ref{fig:simplify}A) can be reduced to a simpler problem, namely: for every target $T$, predict which of the potential regulators $R_i$ regulate $T$ (Figure~\ref{fig:simplify}B). This simplification allowed GENIE3 \cite{huynh-thu_inferringregulatorynetworks_2010} to use Random Forest's \cite{breiman_randomforests_2001} feature importance scores for inferring GRNs. Namely, a Random Forest is trained to predict the expression of a target gene of interest from the expression of potential regulators. The resulting Random Forest inherently allows to extract a feature importance score by observing the effect of each regulator in making a good prediction for the target expression. As in GENIE3, the target expression is first scaled to normalise feature importance scores across different targets.

@@

Computing the case-wise feature importances of a tree consists of the following 8 steps (Figure~\ref{fig:fimp}).
The 'randomness' of a Random Forest is due to only using a subset of the samples in the dataset in order to build a single decision tree. The samples are split into two groups, the 'in-bag' data and the 'out-of-bag' data (Figure~\ref{fig:fimp}A). A decision tree \cite{breiman_classificationregressiontrees_1984} is trained on the in-bag expression of the potential regulators in trying to predict the in-bag target gene expression (Figure~\ref{fig:fimp}B). The target expression of the out-of-bag samples is predicted using the decision tree (Figure~\ref{fig:fimp}C), and the squared error between the real and target expression is computed (Figure~\ref{fig:fimp}D). For each sample in the out-of-bag set, this vector represents how well the decision tree was able to predict the expression of the target gene.

@@

To compute the case-wise GRNs, we implemented the abovementioned methodology in C++ in a modified version of the \texttt{ranger} R/C++ package \cite{wright_rangerfastimplementation_2017}.

@@

\subsection{Clustering of case-wise GRNs}
To perform downstream analysis on the cases, first a $k$-nearest neighbour ($k$NN) graph of the cases is computed.
In order for the $k$NN graph to better emphasise similarities in GRNs rather than absolute euclidean distances, we first reduce the dimensionality of the case-by-interaction matrix to case-by-20 matrix using Landmark Multi-Dimensional Scaling \cite{lee_landmarkmdsensemble_2009} with a Spearman rank distance metric.

Next, KD-trees are used to calculate the $k$NN graph efficiently. The cases in the dataset are visualised and clusted using the Fruchterman-Reingold \cite{fruchterman_graphdrawingforcedirected_1991} and Louvain clustering \cite{blondel_fastunfoldingcommunities_2008}, respectively.

The following R packages provided implementations for each of these algorithms: lmds, RANN, igraph \cite{csardi_igraphsoftwarepackage_2006}.

\subsection{Visualising clustered GRNs}
After Louvain clustering, the interactions of the 50 interactions with highest mean importance per cluster are retained. These interactions are visualised in Cytoscape \cite{shannon_cytoscapesoftwareenvironment_2003}, in which nodes depict genes, edges depict predicted regulatory interactions, coloured according to which cluster they are predicted for. The shape of the arrow denotes the predicted effect of the regulatory interaction.

### 2019-10-16 ch7_incgraph/content.tex
clean up thesis
change cites

\section*{Introduction}
Even the simplest of living organisms already consist of complex biochemical networks which must be able to respond to a variety of stressful conditions in order to survive. 
An organism can be characterised using numerous interaction networks, including gene regulation, metabolic, signalling, and protein-protein interaction. 
The advent of high-throughput profiling methods (e.g. microarrays and RNA sequencing) have allowed to observe the molecular contents of a cell, which in turn have enabled computational network inference methods to reverse engineer the biochemical interaction networks~\cite{albert_networkinferenceanalysis_2007}.
Improving the accuracy of inferred networks has been a long-standing challenge, but the development of ever more sophisticated algorithms and community-wide benchmarking studies have resulted in significant process~\cite{marbach_revealingstrengthsweaknesses_2010, narendra_comprehensiveassessmentmethods_2011, marbach_wisdomcrowdsrobust_2012, aijo_biophysicallymotivatedregulatory_2017}.

Several recent developments involve incorporating topological priors, either to guide the inference process~\cite{lopes_featureselectiontechnique_2014} or post-process the network~\cite{ruyssinck_netterrerankinggene_2016}.
A common prior is that biological networks are highly modular~\cite{rives_modularorganizationcellular_2003}, as they consist of multiple collections of functionally or physically linked molecules~\cite{hartwell_molecularmodularcell_1999,barabasi_networkbiologyunderstanding_2004}. At the lowest level, each module is made up out of biochemical interactions arranged in small topological patterns, which act as fundamental building blocks~\cite{milo_networkmotifssimple_2002}.

Graphlets~\cite{przulj_modelinginteractomescalefree_2004} are one of the tools which could be used to add a topological prior to a biological network,
Graphlets are small connected subnetworks which can be counted to identify which low-level topological patterns are present in a network. By comparing how topologically similar a predicted network is to what would be expected of a true biological network, a predicted network can be optimised in order to obtain a better topology.

By counting the number of occurrences of each of the different graphlets (Fig~\ref{fig_graphlets}A) touching a specific node, one can characterise the topology surrounding it. The graphlet counts of a network can be represented as a matrix with one row for each of the nodes and one column for each of the graphets (Fig~\ref{fig_graphlets}B). 
An orbit represents a class of isomorphic (i.e. resulting in the same structure) positions of nodes within a graphlet (Fig \ref{fig_graphlets}A, coloured in red).
Both graphlets and orbits have been used extensively for predicting the properties of nodes such as protein functionality~\cite{milenkovic_uncoveringbiologicalnetwork_2008,guerrero_characterizationproteasomeinteraction_2008,singh_graphletsignaturebasedscoring_2014} and gene oncogenicity~\cite{milenkovic_optimalnetworkalignment_2010}, by performing network alignment~\cite{kuchaiev_topologicalnetworkalignment_2010,milenkovic_globalnetworkalignment_2013} or using them as a similarity measure in machine learning tasks~\cite{shervashidze_efficientgraphletkernels_2009,vacic_graphletkernelsprediction_2010}.

@@

Eppstein et al. introduced data structures and algorithms for updating the counts of size-three\cite{eppstein_hindexgraphits_2009} and size-four\cite{eppstein_extendeddynamicsubgraph_2012} subgraphs in a dynamic setting. The data structures were determined such that the amortised time is $O(h)$ and $O(h^2)$, respectively, where $h$ is the h-index of the network\cite{hirsch_indexquantifyindividual_2005}.

@@

\subsection*{Timing experiments}
We compared the execution time needed to calculate the graphlet counts in iteratively modified networks between our method and a state-of-the-art non-incremental algorithm, Orca~\cite{hocevar_combinatorialapproachgraphlet_2014}. Orca is a heavily optimised algorithm for counting 5-node graphlets in static networks, and outperforms all other static graphlet counting algorithms by an order of magnitude~\cite{hocevar_combinatorialapproachgraphlet_2014}.

@@

The network models were based on three static network models: Barab\'asi-Albert~\cite{albert_statisticalmechanicscomplex_2002}, Erd\H{o}s--R\'enyi~\cite{erdos_randomgraphs_1959}, and Geometric~\cite{appel_minimumvertexdegree_1997}. These models were adapted to generate evolving networks instead (Figs~\ref{code_ba}, \ref{code_er}, and \ref{code_geo}).
Each model generates an initial network according to the static network model it is based on, and a list of network modifications (removing an edge from or adding an edge to the network). These network modifications are made such that at any given time point in the evolving network, it is likely that the network at its current state could have been generated by the original static network model.

@@

We optimised gene regulatory networks of \textit{E. coli} and \textit{S. cerevisiae}. The predicted networks were generated using the network inference method GENIE3~\cite{huynh-thu_inferringregulatorynetworks_2010} with default parameters. Gene expression data was obtained from COLOMBOS~\cite{moretto_colombosv3leveraging_2016} and GEO~\cite{edgar_geneexpressionomnibus_2002}, respectively. The predicted networks and the optimised versions thereof were compared against respective lists of known gene regulatory interactions~\cite{gama-castro_regulondbversionhighlevel_2016,ma_denovolearninggenomescale_2014}.

@@

Comparing the execution time of IncGraph to the h-index of the networks indicates that the amortised time of IncGraph could be $O(h^3)$ (Fig~\ref{sfig_hindex}). This is in line with the amortised times $O(h)$ and $O(h^2)$ of the algorithm described by Eppstein et al.\cite{eppstein_extendeddynamicsubgraph_2012} for counting three-size and four-size subgraphs respectively.

>>>>

\section*{Introduction}
Even the simplest of living organisms already consist of complex biochemical networks which must be able to respond to a variety of stressful conditions in order to survive. 
An organism can be characterised using numerous interaction networks, including gene regulation, metabolic, signalling, and protein-protein interaction. 
The advent of high-throughput profiling methods (e.g. microarrays and RNA sequencing) have allowed to observe the molecular contents of a cell, which in turn have enabled computational network inference methods to reverse engineer the biochemical interaction networks~ \cite{albert_networkinferenceanalysis_2007}.
Improving the accuracy of inferred networks has been a long-standing challenge, but the development of ever more sophisticated algorithms and community-wide benchmarking studies have resulted in significant process~ \cite{marbach_revealingstrengthsweaknesses_2010, narendra_comprehensiveassessmentmethods_2011, marbach_wisdomcrowdsrobust_2012, aijo_biophysicallymotivatedregulatory_2017}.

Several recent developments involve incorporating topological priors, either to guide the inference process~ \cite{lopes_featureselectiontechnique_2014} or post-process the network~ \cite{ruyssinck_netterrerankinggene_2016}.
A common prior is that biological networks are highly modular~ \cite{rives_modularorganizationcellular_2003}, as they consist of multiple collections of functionally or physically linked molecules~ \cite{hartwell_molecularmodularcell_1999,barabasi_networkbiologyunderstanding_2004}. At the lowest level, each module is made up out of biochemical interactions arranged in small topological patterns, which act as fundamental building blocks~ \cite{milo_networkmotifssimple_2002}.

Graphlets~ \cite{przulj_modelinginteractomescalefree_2004} are one of the tools which could be used to add a topological prior to a biological network,
Graphlets are small connected subnetworks which can be counted to identify which low-level topological patterns are present in a network. By comparing how topologically similar a predicted network is to what would be expected of a true biological network, a predicted network can be optimised in order to obtain a better topology.

By counting the number of occurrences of each of the different graphlets (Fig~\ref{fig_graphlets}A) touching a specific node, one can characterise the topology surrounding it. The graphlet counts of a network can be represented as a matrix with one row for each of the nodes and one column for each of the graphets (Fig~\ref{fig_graphlets}B). 
An orbit represents a class of isomorphic (i.e. resulting in the same structure) positions of nodes within a graphlet (Fig \ref{fig_graphlets}A, coloured in red).
Both graphlets and orbits have been used extensively for predicting the properties of nodes such as protein functionality~ \cite{milenkovic_uncoveringbiologicalnetwork_2008,guerrero_characterizationproteasomeinteraction_2008,singh_graphletsignaturebasedscoring_2014} and gene oncogenicity~ \cite{milenkovic_optimalnetworkalignment_2010}, by performing network alignment~ \cite{kuchaiev_topologicalnetworkalignment_2010,milenkovic_globalnetworkalignment_2013} or using them as a similarity measure in machine learning tasks~ \cite{shervashidze_efficientgraphletkernels_2009,vacic_graphletkernelsprediction_2010}.

@@

Eppstein et al. introduced data structures and algorithms for updating the counts of size-three \cite{eppstein_hindexgraphits_2009} and size-four \cite{eppstein_extendeddynamicsubgraph_2012} subgraphs in a dynamic setting. The data structures were determined such that the amortised time is $O(h)$ and $O(h^2)$, respectively, where $h$ is the h-index of the network \cite{hirsch_indexquantifyindividual_2005}.

@@

\subsection*{Timing experiments}
We compared the execution time needed to calculate the graphlet counts in iteratively modified networks between our method and a state-of-the-art non-incremental algorithm, Orca~ \cite{hocevar_combinatorialapproachgraphlet_2014}. Orca is a heavily optimised algorithm for counting 5-node graphlets in static networks, and outperforms all other static graphlet counting algorithms by an order of magnitude~ \cite{hocevar_combinatorialapproachgraphlet_2014}.

@@

The network models were based on three static network models: Barab\'asi-Albert~ \cite{albert_statisticalmechanicscomplex_2002}, Erd\H{o}s--R\'enyi~ \cite{erdos_randomgraphs_1959}, and Geometric~ \cite{appel_minimumvertexdegree_1997}. These models were adapted to generate evolving networks instead (Figs~\ref{code_ba}, \ref{code_er}, and \ref{code_geo}).
Each model generates an initial network according to the static network model it is based on, and a list of network modifications (removing an edge from or adding an edge to the network). These network modifications are made such that at any given time point in the evolving network, it is likely that the network at its current state could have been generated by the original static network model.

@@

We optimised gene regulatory networks of \textit{E. coli} and \textit{S. cerevisiae}. The predicted networks were generated using the network inference method GENIE3~ \cite{huynh-thu_inferringregulatorynetworks_2010} with default parameters. Gene expression data was obtained from COLOMBOS~ \cite{moretto_colombosv3leveraging_2016} and GEO~ \cite{edgar_geneexpressionomnibus_2002}, respectively. The predicted networks and the optimised versions thereof were compared against respective lists of known gene regulatory interactions~ \cite{gama-castro_regulondbversionhighlevel_2016,ma_denovolearninggenomescale_2014}.

@@

Comparing the execution time of IncGraph to the h-index of the networks indicates that the amortised time of IncGraph could be $O(h^3)$ (Fig~\ref{sfig_hindex}). This is in line with the amortised times $O(h)$ and $O(h^2)$ of the algorithm described by Eppstein et al. \cite{eppstein_extendeddynamicsubgraph_2012} for counting three-size and four-size subgraphs respectively.

### 2019-10-16 ch8_discussion/content.tex
change cites

\texttt{dyngen} has already been successfully used to evaluate trajectory inference\cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory alignment\cite{vandenberge_trajectorybaseddifferentialexpression_2019}, and network inference\cite{pratapa_benchmarkingalgorithmsgene_2019} methods. Furthermore, we showed that it can also be used to evaluate differential network inference methods, and trajectory alignment methods.

@@

Despite \texttt{SCORPIUS} published on bioRxiv in 2016, discussion and usage of the pre-print has largely been limited to comparing the results of a different (linear) TI methods to those of \texttt{SCORPIUS}. Since the pre-print, many improvements have been made to improve the accuracy and robustness of predicted trajectories, to allow scaling up to datasets containing 100'000s of cells, and to providing a more user-friendly interface. Since the beginning of 2019 -- before the publication of our benchmark of TI methods -- several studies have now reported using \texttt{SCORPIUS} as a tool for deriving primary results\cite{damond_maphumantype_2019, cheng_multifactorialheterogeneityvirusspecific_2019, tibbitt_singlecellrnasequencing_2019, hove_singlecellatlasmouse_2019, wouters_singlecellgeneregulatory_2019}.

@@

\section{Optimising regulatory networks}
This project was a spin-off of Netter\cite{ruyssinck_netterrerankinggene_2016}, a tool for reranking GRN predictions using structural network properties. 
Predicted GRNs often are topologically dissimilar to real networks, and as such, Netter quantifies the topological properties of a network and makes incremental changes in order to arrive at a network with a 'better' topological profile.

@@

\textbf{Git}\cite{torvalds_gitfastversion_2005} is a code-revision system that allows multiple users to collaborate on developing code and keep track of what changes were made by whom. 
Since Wouter Saelens and I were often working closely together on a specific part of our software, Git saved us a lot of time by merging together changes developed in parallel. Only in few cases did we need to intervene manually to merge our changes.
Additionally, Github.com allowed us not only to collaborate with each-other, but also correspond and collaborate with other software developers from many different research groups, using Github Issues to discuss problems with other researchers, and Github Pull Requests to contribute code to open-source projects.

Together, we published over 20'000 code contributions (commits) across 50+ software packages\cite{cannoodt_developmentdynverse_2019}. Since many of these packages depend on one another, it is inevitable that changes made to one package would break another package. We wrote many, many unit tests (and still too few), and we let these tests automatically be executed on \textbf{Travis CI}\cite{traviscigmbh_traviscitest_2011}. This way, when we pushed breaking changes to Github, Travis CI would notify that our commit has resulted in one or more unit tests failing. While sometimes it is a hassle to set up, Travis CI has prevented us countless times from generating faulty results due to a faulty underlying function.

Software bugs can be introduced even by incredibly small, seemingly insignificant changes. The R programming language seems particularly susceptible towards getting trapped by its many pitfalls. However, there are many benefits of using R in bioinformatics research context, such as its extensive community of statisticians develop packages for the R ecosystem. The \textbf{tidyverse} packages\cite{wickham_tidytoolsmanifesto_2017}, developed by the RStudio team and many online contributors, completely transformed my experiences with R from tedious struggles to efficient everyday functional programming.

>>>>

\texttt{dyngen} has already been successfully used to evaluate trajectory inference \cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory alignment \cite{vandenberge_trajectorybaseddifferentialexpression_2019}, and network inference \cite{pratapa_benchmarkingalgorithmsgene_2019} methods. Furthermore, we showed that it can also be used to evaluate differential network inference methods, and trajectory alignment methods.

@@

Despite \texttt{SCORPIUS} published on bioRxiv in 2016, discussion and usage of the pre-print has largely been limited to comparing the results of a different (linear) TI methods to those of \texttt{SCORPIUS}. Since the pre-print, many improvements have been made to improve the accuracy and robustness of predicted trajectories, to allow scaling up to datasets containing 100'000s of cells, and to providing a more user-friendly interface. Since the beginning of 2019 -- before the publication of our benchmark of TI methods -- several studies have now reported using \texttt{SCORPIUS} as a tool for deriving primary results \cite{damond_maphumantype_2019, cheng_multifactorialheterogeneityvirusspecific_2019, tibbitt_singlecellrnasequencing_2019, hove_singlecellatlasmouse_2019, wouters_singlecellgeneregulatory_2019, reimegard_combinedmrnaprotein_2019}.

@@

\section{Optimising regulatory networks}
This project was a spin-off of Netter \cite{ruyssinck_netterrerankinggene_2016}, a tool for reranking GRN predictions using structural network properties. 
Predicted GRNs often are topologically dissimilar to real networks, and as such, Netter quantifies the topological properties of a network and makes incremental changes in order to arrive at a network with a 'better' topological profile.

@@

\textbf{Git} \cite{torvalds_gitfastversion_2005} is a code-revision system that allows multiple users to collaborate on developing code and keep track of what changes were made by whom. 
Since Wouter Saelens and I were often working closely together on a specific part of our software, Git saved us a lot of time by merging together changes developed in parallel. Only in few cases did we need to intervene manually to merge our changes.
Additionally, Github.com allowed us not only to collaborate with each-other, but also correspond and collaborate with other software developers from many different research groups, using Github Issues to discuss problems with other researchers, and Github Pull Requests to contribute code to open-source projects.

Together, we published over 20'000 code contributions (commits) across 50+ software packages \cite{cannoodt_developmentdynverse_2019}. Since many of these packages depend on one another, it is inevitable that changes made to one package would break another package. We wrote many, many unit tests (and still too few), and we let these tests automatically be executed on \textbf{Travis CI} \cite{traviscigmbh_traviscitest_2011}. This way, when we pushed breaking changes to Github, Travis CI would notify that our commit has resulted in one or more unit tests failing. While sometimes it is a hassle to set up, Travis CI has prevented us countless times from generating faulty results due to a faulty underlying function.

Software bugs can be introduced even by incredibly small, seemingly insignificant changes. The R programming language seems particularly susceptible towards getting trapped by its many pitfalls. However, there are many benefits of using R in bioinformatics research context, such as its extensive community of statisticians develop packages for the R ecosystem. The \textbf{tidyverse} packages \cite{wickham_tidytoolsmanifesto_2017}, developed by the RStudio team and many online contributors, completely transformed my experiences with R from tedious struggles to efficient everyday functional programming.
